---
title: "Final"
author: "Talula Wilmot"
format: 
  html: 
    toc: true 
    toc-location: right 
editor: visual
---

[My GitHub Repository](https://github.com/talularwilmot/ENVS-193DS_winter-2026_final)

```{r}
#| message: false
#| warning: false
#read in packages from library 
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(tidyverse)
library(dplyr)
library(readr)
#read in data from data folder 
kelp <- read.csv("~/Desktop/ENVS 193DS/git/Final/ENVS-193DS_winter-2026_final/data/Annual_Kelp_All_Years_20250903.csv")
```


# Problem 1. Research Writing 

## a. Transparent statistical methods

In part 1, they used a simple linear regression to determine whether precipitation predicts flooded wetland area (testing the relationship between a continuous predictor and a continuous response).

In part 2, they used an analysis of variance (ANOVA) to test whether the median flooded wetland area differs among the five water year classifications (wet, above normal, below normal, dry, critical drought).

## b. Figure needed 

My coworker could make a box plot to accompany the statement in part 2 by visually showing the audience how similar the **medians** across all five water classifications are. On the x-axis would be water year classification: wet, above normal, below normal, dry, critical drought, and on the y-axis would be flooded wetland area displaying the median as well as quantiles and range. 

## c. Suggestions for rewriting 

Part 1: Flooded wetland area did not show a significant relationship with precipitation. (Linear regression: F() = unknown, r^2^ = unknown, p = 0.11,  $\alpha$ = significance level).

Part 2: Median flooded wetland area appeared similar across the five water year classifications (wet, above normal, below normal, dry, critical drought), indicating that flooding does not differ substantially among these conditions. (ANOVA: F() = unknown, p = 0.12, $\alpha$ = significance level).

## d. Interpretation

One additional variable that could influence the flooded area of wetlands would be management practices on each of the 198 land-owner parcels. This would be a categorical variable listing each parcel as something along the lines of (purposefully flooded, naturally flooded) to delineate if there was artificial flooding during critical breeding periods. 


# Problem 2. Data visualization

## a. cleaning and summarizing 

```{r}
kelp_clean <- kelp %>% 
  clean_names() %>% #change the names of columns to remove spaces and capital letters 
  filter(site %in% c("IVEE", "NAPL", "CARP")) %>% #filter the data to only include these 3 sites
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) %>% #filter out the data from date of 2000-09-22, transect 6 and quadrant 40 
  mutate(site_name = case_when( # adding a column called site_name change the names of sites to be full name 
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) %>% 
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) %>%  # making sure categorical variables are treated as factors and setting level order 
  group_by(site_name, year) %>% #group the data by site_name and year 
  summarize(mean_fronds = mean(fronds)) %>% #calculate the mean number of fronds in each group we just created 
  ungroup() #ungroup the data after calculating the mean 
slice_sample(kelp_clean, n = 5) 
str(kelp_clean)

```
	
## b. Understanding the data 

The number for the value of fronds in this filtered line is -9999, representing a missing datasheet for that transect and quadrant. This information was found in the kelp data sheet on line 73, specified in the notes column. 

## c. Visualiza the data 

```{r}
ggplot(data = kelp_clean, # used clean data 
       aes(x = year, # make x-axis the date
           y = mean_fronds, # make the y axis mean frond count
           color = site_name)) +
  geom_point() +# assign color base
  geom_line() + # create a line between individual points 
  scale_color_manual(values = c("Isla Vista" =  "tan", #color each site specifically 
                                "Naples" = "darkcyan", 
                                "Carpinteria" = "gray")) + 
  labs(x = "Year", # label x-axis 
       y = "Mean kelp fronds per year", # label y -axis 
       color = "Site", # assign color by site
       title = "Sites vary in mean kelp fronds per year") + #Make Title the same as the example
  theme_minimal() + #use theme with blank background 
  theme(text = element_text(family = "times new roman"),
    plot.title = element_text(hjust = 0.5, # make plot title to the left of figure 
                                  size = 10))  # size small enough to fit on page
```

## d. Interpretation 

Isla Vista has the most variable kelp fronds per year showing the highest peaks in the mid 2000's, mid 2010's and around 2021, while also maintaining similar lows across all 3 sites showing high fluctuation. Carpinteria tends to have the least variable fronds per year remaining mostly in between the peaks and lows of Naples and Isla Vista with some lower periods around 2020 but no incredibly high years. 

# Problem 3. Data Analysis 

## a. Comparing visualizations 

## b. Designing an analysis 

## c. Check any assumptions and run an analysis 

## d. Create a visualization 

## e. Write a caption 

## f. Write about your results 

## g. Sharing affective visualization 

In class! 







