---
title: "Final"
author: "Talula Wilmot"
format: 
  html: 
    toc: true 
    toc-location: right 
editor: visual
---

[My GitHub Repository](https://github.com/talularwilmot/ENVS-193DS_winter-2026_final)

```{r}
#| message: false
#| warning: false
#read in packages from library 
library(tidyverse)
library(here)
library(janitor)
library(readxl)
library(tidyverse)
library(dplyr)
library(readr)
library(ggeffects)
library(MuMIn)
library(DHARMa)
#read in data from data folder 
kelp <- read.csv("~/Desktop/ENVS 193DS/git/Final/ENVS-193DS_winter-2026_final/data/Annual_Kelp_All_Years_20250903.csv")
Personal <- read_csv("~/Desktop/ENVS 193DS/git/Final/ENVS-193DS_winter-2026_final/data/My Data ES 193 - Sheet1.csv")
mopl <- read_csv("~/Desktop/ENVS 193DS/git/Final/ENVS-193DS_winter-2026_final/data/MOPL_nest-site_selection_Duchardt_et_al._2019.csv")
```


# Problem 1. Research Writing 

## a. Transparent statistical methods

In part 1, they used a simple linear regression to determine whether precipitation predicts flooded wetland area (testing the relationship between a continuous predictor and a continuous response).

In part 2, they used an analysis of variance (ANOVA) to test whether the median flooded wetland area differs among the five water year classifications (wet, above normal, below normal, dry, critical drought).

## b. Figure needed 

My coworker could make a box plot to accompany the statement in part 2 by visually showing the audience how similar the **medians** across all five water classifications are. On the x-axis would be water year classification: wet, above normal, below normal, dry, critical drought, and on the y-axis would be flooded wetland area displaying the median as well as quantiles and range. 

## c. Suggestions for rewriting 

Part 1: Flooded wetland area did not show a significant relationship with precipitation. (Linear regression: F() = unknown, r^2^ = unknown, p = 0.11,  $\alpha$ = significance level).

Part 2: Median flooded wetland area appeared similar across the five water year classifications (wet, above normal, below normal, dry, critical drought), indicating that flooding does not differ substantially among these conditions. (ANOVA: F() = unknown, p = 0.12, $\alpha$ = significance level).

## d. Interpretation

One additional variable that could influence the flooded area of wetlands would be management practices on each of the 198 land-owner parcels. This would be a categorical variable listing each parcel as something along the lines of (purposefully flooded, naturally flooded) to delineate if there was artificial flooding during critical breeding periods. 


# Problem 2. Data visualization

## a. cleaning and summarizing 

```{r}
kelp_clean <- kelp %>% 
  clean_names() %>% #change the names of columns to remove spaces and capital letters 
  filter(site %in% c("IVEE", "NAPL", "CARP")) %>% #filter the data to only include these 3 sites
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) %>% #filter out the data from date of 2000-09-22, transect 6 and quadrant 40 
  mutate(site_name = case_when( # adding a column called site_name change the names of sites to be full name 
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
)) %>% 
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) %>%  # making sure categorical variables are treated as factors and setting level order 
  group_by(site_name, year) %>% #group the data by site_name and year 
  summarize(mean_fronds = mean(fronds)) %>% #calculate the mean number of fronds in each group we just created 
  ungroup() #ungroup the data after calculating the mean 
slice_sample(kelp_clean, n = 5) 
str(kelp_clean)

```
	
## b. Understanding the data 

The number for the value of fronds in this filtered line is -9999, representing a missing datasheet for that transect and quadrant. This information was found in the kelp data sheet on line 73, specified in the notes column. 

## c. Visualiza the data 

```{r}
ggplot(data = kelp_clean, # used clean data 
       aes(x = year, # make x-axis the date
           y = mean_fronds, # make the y axis mean frond count
           color = site_name)) +
  geom_point() +# assign color base
  geom_line() + # create a line between individual points 
  scale_color_manual(values = c("Isla Vista" =  "tan", #color each site specifically 
                                "Naples" = "darkcyan", 
                                "Carpinteria" = "gray")) + 
  labs(x = "Year", # label x-axis 
       y = "Mean kelp fronds per year", # label y -axis 
       color = "Site", # assign color by site
       title = "Sites vary in mean kelp fronds per year") + #Make Title the same as the example
  theme_minimal() + #use theme with blank background 
  theme(text = element_text(family = "times new roman"),
    plot.title = element_text(hjust = 0.5, # make plot title to the left of figure 
                                  size = 10))  # size small enough to fit on page
```

## d. Interpretation 

Isla Vista has the most variable kelp fronds per year showing the highest peaks in the mid 2000's, mid 2010's and around 2021, while also maintaining similar lows across all 3 sites showing high fluctuation. Carpinteria tends to have the least variable fronds per year remaining mostly in between the peaks and lows of Naples and Isla Vista with some lower periods around 2020 but no incredibly high years. 

# Problem 3. Data Analysis 

## a. Response variable

In this data set the 1's and 0's represent nest site occupancy by Mountain plover. A 1 means that there was a nest present and a 0 means that there was no nest present. 

## b. Purpose of study  

Visual obstruction is a continuous variable that measures vegetation structure (height and density) around a potential nest site. As visual obstruction increases, the probability of Mountain Plover nest site use would likely decrease because plovers prefer areas with short, sparse vegetation and more open ground so they can detect predators and maintain visibility around the nest. This information was found in the Methods section under Adult mountain plover density - Data collection, paragraph 3. 

Distance to the prairie dog colony edge is also a continuous variable representing how far a potential nest site is from the boundary of a prairie dog colony. As distance from the colony edge increases, the probability of nest site use may decrease, prairie dogs clip vegetation and disturb soil allowing for greater visibility of predators. This information was found in the Methods section in Table 1. 

## c. Table of models 
Model Number| Visual Obstruction | Distance to colony | Model Description 

0           |                    |                    | null model 

1           |       X            |        X           | saturated model

2           |       X            |                    | visual obstruction only

3           |                    |        X           | colony dist. only 

## d. Run the models 

```{r}
mopl_clean <- mopl %>%  
  clean_names() %>%  # clean the names of the columns to remove capitals and spaces 
   mutate(used_bin = ifelse(id == "used", 1, 0)) %>%  # convert id to 1/0
  select(used_bin, edgedist, vo_nest) #keep only these columns 
# model 0: null model
model0 <- glm(
  used_bin ~ 1, 
  data = mopl_clean,
   family = binomial
)

# model 1: all predictors

model1 <- glm(
  used_bin ~ edgedist + vo_nest, 
  data = mopl_clean,
   family = binomial
)
# model 2: visual obstruction only 

model2 <- glm(
  used_bin ~ vo_nest,  
  data = mopl_clean, 
  family = binomial
)

# model 3: Colony distance only 

model3 <- glm(
  used_bin ~ edgedist,  
  data = mopl_clean,
   family = binomial
)
```

## e. Select the best model 

```{r}
AICc(model0, 
     model1, 
     model2, 
     model3) %>% 
  arrange(AICc)
```
The best model as determined by Akaike’s Information Criterion (AIC) was Model 2 using visual obstruction to explain the probability of a nesting site. 

## f. Check the diagnostics 

```{r}
plot(simulateResiduals(model2)) #run diagnostic checks 
```

## g. Visualize the model predictions 

```{r}
model2_preds <- ggpredict(model2, terms = "vo_nest")
ggplot(data = mopl_clean, 
       aes(x = vo_nest, 
           y = used_bin)) + 
  geom_point(color = "darkcyan") + 
  geom_ribbon(data = model2_preds, 
              aes(x = x, 
                  y = predicted, 
                  ymin = conf.low, 
                  ymax = conf.high),
                  alpha = 0.1, 
              color = "orange") + 
  geom_line(data = model2_preds, 
            aes(x = x, 
                y = predicted), 
            linewidth = 1) + 
  scale_y_continuous( limits = c(0,1), 
                      breaks = c(0,1)) + 
  labs(x = "Visual obstruction (height and density)", 
       y = "Nest site used (0 = no, 1 = yes)") + 
  theme_minimal()
```

## h. Write a caption 
**Figure 1. Binomial regression of suitable habitat based on obstruction** Blue points represent raw data points of site occupancy or vaccancy for Mountain Plovers. Line represents predicted equation for suitable habitat while orange lines outline the 95% confidence interval for those estimates. Data was collected by Duchardt, Courtney; Beck, Jeffrey; Augustine, David (2020). Data from: Mountain Plover habitat selection and nest survival in relation to weather variability and spatial attributes of Black-tailed Prairie Dog disturbance [Dataset]. Dryad. https://doi.org/10.5061/dryad.ttdz08kt7 and accessed via csv download. 


## i. Calculate model predictions 

```{r}
gtsummary::tbl_regression(model2, exponentiate = TRUE)
```


```{r}
ggpredict(model2, terms = "vo_nest [0]") # use model 2 and gg predict to predict the probability of occupancy at vo index 0 
```

## j. Interpret your results 

Mountain plovers tend to chose nest sights with no visual obstruction because they have a better peripheral view of potential predators. Prairie dog colonies often create lack of obstruction from soil disturbance and consumption of nearby shrubs but AIC was lower for obstruction alone. With a visual obstruction with an index of 0 predicted nest sight occupancy was 0.73 (95% CI: 0.65, 0.80). For every 1 unit increase in visual obstruction the odds of nest side occupancy decrease by a factor of 0.58, (95% CI: 0.47, 0.69, p< 0.001, $\alpha$ = 0.05)

# Problem 4. Affective and exploratory visualizations 

## a. Comparing visualizations 

How are the visualizations different from each other in the way you have represented your data?

In my exploratory analysis I only compared 1 variable at a time, however in my affective visualization I chose to represent a multitude of variables. 

What similarities do you see between all your visualizations?

I see a similar trend between all my visuals in my data showing a kind of scattered distribution with a possible linear trend, as mileage increases. 

What patterns (e.g. differences in means/counts/proportions/medians, trends through time, relationships between variables) do you see in each visualization? Are these different between visualizations? If so, why? If not, why not?

I see a possible parabolic trend of my data through the distances that I ran. From 0- 4 miles my paces seemed to increase but then above that I seem to have possibly rounded a bend and gotten quicker. However It could also potentially be a linear trend with some outliers as I have only done a few longer distance runs. This trend is shown both in my exploratory and affective visualization, but my other exploratory visualization was based off of a categorical variable and it doesn't show any explicit trends other than running tended to be the most frequent workout type. 

What kinds of feedback did you get during week 9 in workshop or from the instructors? How did you implement or try those suggestions? If you tried and kept those suggestions, explain how and why; if not, explain why not.

During week 9 the feedback I got from my peer reviewers was to increase detail on the foreground of my visualization, potentially using the location where I always run as a sort of nature landscape, and I implemented those suggestions into my painting. I did not receive any feedback on including additional variables or really structurally changing the work on my index card feedbacks. 

## b. Designing an analysis 

One analysis that I could apply to my research question is a linear regression. This is appropriate for my question because I would assume that especially as a beginner runner my pace would slow with the addition of miles and I am interested to test if that assumption is true. Additionally both of my variables are continuous. 

## c. Check any assumptions and run an analysis 

### run analysis 

```{r}
personal_clean <- clean_names(Personal) #clean the names of personal data to only include lower case and remove spaces 
running_data <- personal_clean %>% #use to personal_clean data and filter to only include rows with the workout type equal to running 
  filter(workout_type == "Running") #filter the data to only include running observations
running_model <- lm(as.numeric(pace) ~ distance_mi, #use pace as dependent and milage as predictor
                    data = running_data) #select from filtered data set 
summary(running_model) #display summary 
```

### check assumptions 

```{r}
par(mfrow = c(2,2)) #show plots in 2 by 2 grid 
plot(running_model) #create diagnostic plots for our running model
```

## d. Create a visualization 
```{r}
running_preds <- ggpredict(running_model, terms = "distance_mi")
ggplot(data = running_data, 
       aes(x = distance_mi, 
           y = pace)) + 
  geom_point(color = "darkcyan") + 
  geom_ribbon(data = running_preds, 
              aes(x = x, 
                  y = predicted, 
                  ymin = conf.low, 
                  ymax = conf.high),
                  alpha = 0.1, 
              color = "orange") + 
  geom_line(data = running_preds, 
            aes(x = x, 
                y = predicted), 
            linewidth = 1) + 
  labs(x = "Distance (mi)", 
       y = "Pace (hr:mm:ss)") + 
  theme_minimal()
```

## e. Write a caption 

**Figure 2. Linear regression of running vs. pace with 95% confidence interval** Blue dots represent underlying data of average pace for duration of run. Orange lines outlining ribbon represent the upper and lower confidence interval for predicted linear relationship. Black line represents predicted linear equation. Data was collected by Author and access via CSV download. 

## f. Write about your results 

The pace increased as distance increased (linear regression: F(1,17) = 8.573, R^2^ = 0.3, p = 0.009, $\alpha$ = 0.05) Based on model predictions for every 1 unit increase in distance we expect to see an increase in pace of 9.6 seconds +/- 3.3 seconds. While the model did show significance the R^2^ being 0.3 means that not much of the variation in the model was explained by a linear regression. 

## g. Sharing affective visualization 

In class! 







